A Guide to Designing Experiments to Test Statistical Graphics

### 2025-03-17 index.tex
diff files

\title{A Guide to Designing Experiments to Test Statistical Graphics}
\author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
\date{2025-03-14}

@@

It is extremely challenging to design studies which strike the right
balance between experimental control (i.e.~internal validity) and
generalizability to a wider context (i.e.~external validity). Simply
asking people to read quantities off of a graph may not generalize
beyond the questions asked or the data used in the chart (Croxton, 1932;
Croxton \& Stryker, 1927; Eells, 1926; Huhn, 1927), but designing a
study that is sufficiently robust to those issues requires manipulation
or control of so many factors that the amount of participants and trials
quickly becomes daunting or unaffordable. In this paper, we attempt to
distill the experience gained from conducting several different types of
graphics experiments (Hofmann et al., 2012; Robinson, 2022; VanderPlas
et al., 2019; Vanderplas et al., 2024; VanderPlas \& Hofmann, 2015,
2017), discussing the use of different testing methods (Vanderplas et
al., 2020), the process of designing a graphical experiment, and
analysis of the resulting empirical data. It is our hope that this paper
will lower the barriers that exist for conducting empirical graphics
research and reduce the probability of costly mistakes.

Section~\ref{sec-testing-methods} discusses different methods for
testing graphics, and which methods best address different levels of
user engagement. In Section~\ref{sec-model-dev}, we discuss the process
of developing a data-generating model used to control the statistical
features of data in the tested visualizations. Model development is a
nuanced and iterative process that ultimately determines the success and
generalizability of the experimental results. We explore different
experimental design considerations in Section~\ref{sec-exp-design} and
then move to the importance of pilot testing in
Section~\ref{sec-pilot-test}. Finally, we provide some common analysis
strategies in Section~\ref{sec-analysis}, including strategies for
handling the unexpected data features which are so common in graphical
testing experiments.

@@

It is important to consider the level of user engagement which is
necessary to complete a particular visual or graphical task. For
instance, testing whether someone can detect an effect such as a linear
trend in noisy data is a perceptual question. Perceptual questions are
often examined experimentally using methods which allow the user to
interact with the data on a basic visual level: users are presented with
a visual stimulus and answer yes/no questions to indicate whether the
effect is detected. Numerical estimation is another common task when
testing graphics: in these experiments, the participant views a chart,
estimates the requested numerical quantity, and enters the estimate into
the application through a numerical input, slider, or other form
element. Sometimes, it is possible to set up a scenario where the user
adjusts the plot using a set of controls designed to provide a fixed set
of interactive operations. This type of user engagement was used to
assess the strength of the sine illusion (VanderPlas \& Hofmann, 2015):
users adjusted the strength of a transformation designed to correct the
illusion until the lines appeared to be the same length, as shown in
Figure~\ref{fig-sine-illusion}, providing a direct measure of the
magnitude of the sine illusion's effect. In other situations, it may be
preferable to have the user directly interact with the visual stimulus.
In Vanderplas et al. (2024), participants were asked to rotate and
interact with a 3D rendered bar chart; the application recorded user
interactions and corresponding rotation matrices, providing insight into
the visual comparisons the user may have been performing. This
information was used as a supplement to the explicitly provided
estimates, providing some contextual information as well as the ability
to identify the level of participant engagement with the questions. When
experiments are conducted as part of classroom experiential learning, it
is sometimes helpful to be able to separate the low-effort participants
from those who were fully intellectually engaged in the task.
Interactive graphics provide another level of user engagement that can
be much more open-ended. With interactive graphics, researchers can ask
participants to directly annotate plots, toggle aesthetics, and
highlight groups and plot features. Careful implementation of the
experiment application may allow for each of these interactions to be
recorded and analyzed, producing a rich, if messy, set of data that may
allow researchers to tease apart visual estimation error from common
shortcuts such as rounding used during direct numerical estimation.

@@

Visual inference (Buja et al., 2009; Wickham et al., 2010) is another
useful testing tool for perceptual questions such as ``which chart
displays this data more clearly'' (Hofmann et al., 2012) while
simultaneously assessing the statistical significance of the graphical
finding in a chart. Visual inference charts are often called ``lineups''
in analogy to the criminal procedure where the suspect is placed in a
line with several other individuals with similar characteristics. In a
graphical lineup procedure, there is a target plot containing the real
data, embedded in an array of innocent ``null'' plots generated through
resampling or simulation. If viewers consistently pick the target plot
at a higher rate than any of the null plots, the target plot is said to
be visually significant (Loy \& Hofmann, 2013; Majumder et al., 2013)
and a ``see'' value, the visual analogue of a \(p\)-value (Chowdhury et
al., 2020), can be calculated using the \texttt{vinference} R package or
the process described in VanderPlas et al. (2021). The details of this
calculation are beyond the scope of this broader discussion of how to
test charts, but more detail on visual inference is provided in
\textless insert citation to visual inference WIRE article under
development\textgreater.

In another variation of the statistical lineup procedure, data generated
from two models are compared, with target plots from each model embedded
in the array of \(K\) total plots. The \(K-2\) null plots are
constructed from a mixture model that blends the two competing models
(VanderPlas \& Hofmann, 2017). Viewers are asked to select the panel(s)
which are most different, and the primary source of information are
trials in which viewers selected the target from one model but not the
other, indicating that the display method used allowed viewers to
differentiate one model's data (but not the other) from the nulls
created through a mixture model. This variation allows the experimenter
to assess graphical design choices to determine whether they effectively
emphasize structural differences in the data (VanderPlas \& Hofmann,
2017).

@@

To assess the viewer's \emph{understanding} of information shown in a
chart, we must ask questions and allow the user to provide feedback.
User feedback may be collected on a numerical scale or through the use
of written comments, recorded ``think-aloud'' processes, and other more
qualitative interaction methods. In some studies, asking users to
interpret a chart within a larger scenario can be effective, as in
Figure~\ref{fig-estimation-describe}, while in others it is more helpful
to ask users to explain answers. In visual inference studies, asking
users why a specific panel was chosen has been demonstrated to provide
rich insight into otherwise confusing numerical results (VanderPlas \&
Hofmann, 2017).

@@

Even when great care is taken with the construction of the question,
participant answer accuracy is fundamentally limited by the fact that
many participants do not read and interpret the question with the care
and precision that it was written. Questions that ask participants to
e.g.~estimate the multiplicative change in a quantity at two time points
may be misunderstood as asking for an estimate of the additive
difference, and the resulting estimates are then one or more orders of
magnitude off of the correct answer. This is one area where lineup
methods are convenient - they do not depend on participants to
understand the nuances of language or scenarios built around the chart
under investigation. However, in some situations it may be sufficient to
ask participants to estimate direct numerical quantities that have
little contextual information, as done in (VanderPlas et al., 2019) when
assessing the accuracy of framed plots re-created from the Statistical
Atlas.

@@

@@

Once the graphical task has been identified, it is necessary to develop
a model which can be used to explore the graphical features of interest
in a precise manner. This is the single longest part of the entire
experimental design and execution process, in part because choosing a
model that replicates important visual features of the data is extremely
complex (Cook et al., 2021; Hullman \& Gelman, 2021; VanderPlas, 2021).

There are two main options when developing a statistical model for
graphical testing: start with a large data set and sample from that data
set (Hofmann et al., 2012), or start from a model and sample data from
that model generating process (Robinson, 2022; VanderPlas \& Hofmann,
2015, 2017). This decision is largely determined by the availability of
a large data set containing the requisite features of interest and the
qualities being manipulated in the experiment. For instance, Hofmann et
al. (2012) used samples of different sizes from a pre-existing data set
to manipulate the amount of signal in each comparison; with a small
sample, there is less signal and the same amount of noise, making the
true plot harder to spot. In many situations, though, a convenient data
set with the right properties is harder to acquire, and it becomes
necessary to develop a sampling model to generate data for user
evaluation.

@@

The choice of the parameter space used in testing is crucial to gain
insight from a study without putting too much burden on participants
with overlong studies. Choosing an appropriate space for testing
parameters is a well-known problem in psychometric testing: the space
considered should cover the area between `only some activation' to
`almost full activation' of an appropriate psychometric function (Schütt
et al., 2016; Valentin et al., 2024). When testing charts, visual
assessment is obviously key, but researchers can make use of statistical
indices related to the testing condition to narrow the parameter space
to a reasonable and efficient subset from which maximal information can
be acquired.\\
These statistical indices may also serve as quantitative proxies for the
difficulty of the visual task.

To identify a statistical proxy for visual difficulty that may help with
narrowing the parameter space, it can be useful to consider numerical
measures used to estimate the same types of visual information that will
be assessed in the experiment. For instance, we have used

@@

Once an appropriate set of parameters are identified using the numerical
screening method, it is important to calibrate these parameter
selections visually. No numerical statistic is a perfect measure of what
we actually see: at best, they are approximations of what we might
potentially see. We have found it to be useful to have one experimenter
calibrate the model parameters at a gross level, and then have another
experimenter narrow in on the parameters which are visually reasonable
within the selected range. Then, both examiners visually inspect a large
number of plots generated using those parameters to get a sense for how
difficult the task at hand is (this strategy is also described by Lu et
al. (2022)). At some point, all experimenters become so visually
saturated with the nuances of the data generating mechanism that it may
become necessary to ``sanity check'' the protocol with family members,
friends, and colleagues. These informal surveys provide extremely useful
feedback and can help to counteract the visual saturation of being
immersed in the design of a visualization experiment for months at a
time.

@@

We cannot overstate the importance of visual assessment of your model
stimuli, preferably with fresh eyes. We highly recommend performing
several rounds of think-aloud pilot testing before deploying an
experiment. In support of this assessment, we offer up a cautionary tale
of our own experience: that of VanderPlas \& Hofmann (2017), where we
designed an experiment to test which plot aesthetics promoted discovery
of linear trends and/or clusters.

The experiment was a full 2x3x3 factorial exploration of three data
generating parameters, with 3 replicates at each parameter combination
(54 data sets) and 10 aesthetic combinations (for a total of 540
lineups). Each lineup had 20 different sub-panels, so we should have
carefully visually inspected some 10,800 different panels. As is evident
from the fact that we're telling this story as a cautionary tale, we
missed a critical problem with our data-generating mechanism: when
clusters were assigned to randomly generated data after the fact, we
didn't control the cluster size, leading to clusters of one or two
points in relatively few sub-panels. This became particularly noticeable
when bounding ellipses were added to the plot, as the method used to
generate those ellipses required at least 3 points in the cluster. The
missing boundary ellipse in the corresponding sub-panels escaped our
notice during the stimuli proof-reading phase of the experiment, but did
not escape the notice of our participants, who only needed to examine
about 10 lineups each (around 200 panels). An example of one of the
problematic lineups is shown in Figure~\ref{fig-lineup-problems}: many
participants selected panel 16 because of the missing ellipse; not a
wrong choice, but certainly not the effect we intended to test.

@@

\section{Experimental Design Considerations}\label{sec-exp-design}

It would be difficult to develop a full data generating model without
some idea of the experimental design: the basic structure of the
parameters which are to be manipulated, how the users will be tested,
and so on. These experimental design factors are fairly natural for
scientists to accumulate over the course of imagining and planning an
experiment. When conducting graphical tests, however, there are
additional considerations beyond those taught in a standard experimental
design course.

\subsection{Cognitive Load}\label{cognitive-load}

The primary human design factor to be aware of is that visual tasks and
assessing statistical graphics can be extremely cognitively taxing. In
our experience, it is difficult to expect participants to evaluate more
than about 15 charts in one sitting. If users are asked to deeply engage
and answer multiple questions about each chart, this limit may be lower
(8-10), but even with a relatively simple task, as in lineup methods, it
is hard for participants to accurately evaluate more than about 15
charts. Tasks which are more interactive, such as `You Draw It', may be
somewhat easier for participants, but it is unlikely that participants
would be willing to complete more than about 20 tasks in one sitting
even with tasks that require fewer decisions and more engagement. At
some point, participants' interest in accuracy will decline, and the
researcher is better off using a larger number of participants
completing fewer tasks within the window where participants are
attentive and engaged.

@@

It can be extremely helpful to include ``practice'' demonstrations of
the task to show the basic process, logic, and reasoning. While it is
tempting to make these tasks fully representative of the type of
judgement which will be required of participants, practice tasks which
are too close to the experimental task may bias participants; we have
found that it works well to have a relatively easy practice task which
utilizes a slightly different type of plot and/or type of data than what
will be tested in the experiment. In cases which require interactivity,
gif animations of the task being carried out are useful, as are
additional visual cues, such as the yellow box used in the `You Draw It'
task\footnote{See a gif of testing with `You Draw It'
  \href{https://i.imgur.com/GM5YSen.gif}{here}} to indicate that there
were points which were not completed.

\subsection{Accounting for Participant
Variability}\label{accounting-for-participant-variability}

While studies have found some relationship between lineup performance
and demographic factors (VanderPlas \& Hofmann, 2016), these differences
are relatively small when participants are recruited from online testing
platforms like Amazon Mechanical Turk or Prolific, or when participants
are recruited from a university student population. Studies have also
not found a strong relationship between visual task performance and
participant recruitment method (VanderPlas et al., 2019), perhaps in
part because demographics which use social media sites and demographics
completing online research tasks for pay overlap heavily. However, when
participants are recruited using a statistically representative sample
of the wider population, education becomes an extremely useful predictor
of ability to effectively read and draw conclusions about a chart (Rice
et al., 2024).\\
In almost every situation, though, we find that some participants are
very good at evaluating graphics and some participants are not; as a
result, it can be extremely helpful to use random effects models with an
effect for individual participants.

>>>>

\title{A Guide to Designing Experiments to Test Statistical Graphics}
\author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
\date{2025-03-16}

@@

It is extremely challenging to design studies which strike the right
balance between experimental control (i.e.~internal validity) and
generalizability to a wider context (i.e.~external validity). Simply
asking people to read quantities off of a graph may not generalize
beyond the questions asked or the data used in the chart (Croxton, 1932;
Croxton \& Stryker, 1927; Eells, 1926; Huhn, 1927), but designing a
study that is sufficiently robust to those issues requires manipulation
or control of so many factors that the amount of participants and trials
quickly becomes daunting or unaffordable. In addition, when conducting
graphics experiments, researchers are in the unusual position of being
both the subject matter expert and the statistician, providing an
unusual amount of control over not just the experimental design but also
the specific treatments, levels, and experimental protocols. The amount
of choices required to develop, pilot, and run an experiment can be
overwhelming. In this paper, we attempt to distill the experience gained
from conducting several different types of graphics experiments (Hofmann
et al., 2012; Robinson, 2022; Vanderplas et al., 2019, 2024; Vanderplas
\& Hofmann, 2015, 2017), discussing the use of different testing methods
(Vanderplas et al., 2020), the process of designing a graphical
experiment, and analysis of the resulting empirical data. It is our hope
that this paper will lower the barriers that exist for conducting
empirical graphics research and reduce the probability of costly
mistakes.

Section~\ref{sec-testing-methods} discusses different methods for
testing graphics, and which methods best address different levels of
user engagement. In Section~\ref{sec-model-dev}, we discuss the process
of developing the data-generating model used to control the statistical
features of data in the tested visualizations. Model development is a
nuanced and iterative process that ultimately determines the success and
generalizability of the experimental results. In
Section~\ref{sec-exp-dev}, we discuss the design of the experimental
protocol - the choice of platform, number and type of trials, and flow
of the experiment. We briefly consider different experimental design
considerations in Section~\ref{sec-exp-design}, but focus primarily on
factors specific to graphics experiments, and then move to the
importance of pilot testing in Section~\ref{sec-pilot-test}. Finally, we
provide some common analysis strategies in Section~\ref{sec-analysis},
including strategies for handling the unexpected data features which are
so common in graphical testing experiments.

@@

It is important to consider the level of user engagement which is
necessary to complete a particular visual or graphical task. For
instance, testing whether someone can detect an effect such as a linear
trend in noisy data is a perceptual question. Perceptual questions are
often examined experimentally using methods which allow the user to
interact with the data on a basic visual level: users are presented with
a visual stimulus and answer yes/no questions to indicate whether the
effect is detected. Numerical estimation is another common task when
testing graphics: in these experiments, the participant views a chart,
estimates the requested numerical quantity, and enters the estimate into
the application through a numerical input, slider, or other form
element. Sometimes, it is possible to set up a scenario where the user
adjusts the plot using a set of controls designed to provide a fixed set
of interactive operations. This type of user engagement was used to
assess the strength of the sine illusion (Vanderplas \& Hofmann, 2015):
users adjusted the strength of a transformation designed to correct the
illusion until the lines appeared to be the same length, as shown in
Figure~\ref{fig-sine-illusion}, providing a direct measure of the
magnitude of the sine illusion's effect. In other situations, it may be
preferable to have the user directly interact with the visual stimulus.
In Vanderplas et al. (2024), participants were asked to rotate and
interact with a 3D rendered bar chart; the application recorded user
interactions and corresponding rotation matrices, providing insight into
the visual comparisons the user may have been performing. This
information was used as a supplement to the explicitly provided
estimates, providing some contextual information as well as the ability
to identify the level of participant engagement with the questions. When
experiments are conducted as part of classroom experiential learning, it
is sometimes helpful to be able to separate the low-effort participants
from those who were fully intellectually engaged in the task.
Interactive graphics provide another level of user engagement that can
be much more open-ended. With interactive graphics, researchers can ask
participants to directly annotate plots, toggle aesthetics, and
highlight groups and plot features. Careful implementation of the
experiment application may allow for each of these interactions to be
recorded and analyzed, producing a rich, if messy, set of data that may
allow researchers to tease apart visual estimation error from common
shortcuts such as rounding used during direct numerical estimation.

@@

Visual inference (Buja et al., 2009; Wickham et al., 2010) is another
useful testing tool for perceptual questions such as ``which chart
displays this data more clearly'' (Hofmann et al., 2012) while
simultaneously assessing the statistical significance of the graphical
finding in a chart. Visual inference charts are often called ``lineups''
in analogy to the criminal procedure where the suspect is placed in a
line with several other individuals with similar characteristics. In a
graphical lineup procedure, there is a target plot containing the real
data, embedded in an array of innocent ``null'' plots generated through
resampling or simulation. If viewers consistently pick the target plot
at a higher rate than any of the null plots, the target plot is said to
be visually significant (Loy \& Hofmann, 2013; Majumder et al., 2013)
and a ``see'' value, the visual analogue of a \(p\)-value (Chowdhury et
al., 2020), can be calculated using the \texttt{vinference} R package or
the process described in Vanderplas et al. (2021). The details of this
calculation are beyond the scope of this broader discussion of how to
test charts, but more detail on visual inference is provided in
\textless insert citation to visual inference WIRE article under
development\textgreater.

In another variation of the statistical lineup procedure, data generated
from two models are compared, with target plots from each model embedded
in the array of \(K\) total plots. The \(K-2\) null plots are
constructed from a mixture model that blends the two competing models
(Vanderplas \& Hofmann, 2017). Viewers are asked to select the panel(s)
which are most different, and the primary source of information are
trials in which viewers selected the target from one model but not the
other, indicating that the display method used allowed viewers to
differentiate one model's data (but not the other) from the nulls
created through a mixture model. This variation allows the experimenter
to assess graphical design choices to determine whether they effectively
emphasize structural differences in the data (Vanderplas \& Hofmann,
2017).

@@

To assess the viewer's \emph{understanding} of information shown in a
chart, we must ask questions and allow the user to provide feedback.
User feedback may be collected on a numerical scale or through the use
of written comments, recorded ``think-aloud'' processes, and other more
qualitative interaction methods. In some studies, asking users to
interpret a chart within a larger scenario can be effective, as in
Figure~\ref{fig-estimation-describe}, while in others it is more helpful
to ask users to explain answers. In visual inference studies, asking
users why a specific panel was chosen has been demonstrated to provide
rich insight into otherwise confusing numerical results (Vanderplas \&
Hofmann, 2017).

@@

Even when great care is taken with the construction of the question,
participant answer accuracy is fundamentally limited by the fact that
many participants do not read and interpret the question with the care
and precision that it was written. Questions that ask participants to
e.g.~estimate the multiplicative change in a quantity at two time points
may be misunderstood as asking for an estimate of the additive
difference, and the resulting estimates are then one or more orders of
magnitude off of the correct answer. This is one area where lineup
methods are convenient - they do not depend on participants to
understand the nuances of language or scenarios built around the chart
under investigation. However, in some situations it may be sufficient to
ask participants to estimate direct numerical quantities that have
little contextual information, as done in (Vanderplas et al., 2019) when
assessing the accuracy of framed plots re-created from the Statistical
Atlas.

@@

\section{Experiment Development Life
Cycle}\label{experiment-development-life-cycle}

Developing a graphics experiment is often a highly iterative process,
but it can help to approach the design process by first optimizing the
model and data generation method before spending time on optimizing the
specific stimuli or customizing the data collection platform. This is
important because the model parameters and data generation process
inform the experiment structure and thus impact decisions made
downstream.

Once the model and data generating mechanism are set, it is useful to
revisit the primary questions of interest and determine how to measure
the responses effectively. Secondary measures, such as response time,
free responses, and confidence level should also be determined. These
choices will inform the choice of a data collection platform and may
also inform the participant recruitment method.

Next, we recommend developing a preliminary data analysis plan,
specifying the general category of model which will be used
(e.g.~generalized linear mixed-effects model, t-test) and the contrasts
which are most interesting. This sets up the experimental design
decisions, but also ensures that as the data collection platform and
process is developed, any design constraints are considered. Development
of the data collection application is the next step, using draft
graphics and the set of participant response measures of interest.

There are at least 3 stages of testing in a graphics experiment:
informal tests, a pilot study, and the main experiment. The informal
tests are critical for identifying issues with the data collection
application, but can also be used to calibrate the number of tasks
required of each participant. As the number and complexity of tasks
increases, the number of trials we can ask participants to complete
during a session decreases. The informal testing stage allows
researchers to consider the tradeoffs inherent in the decision to reduce
the amount of information collected for each task, reduce the number of
tasks, or mitigate participant fatigue in other ways.

During preliminary testing, we use an optimistic number of trials per
participant, so that we can determine when participants become overly
fatigued. For instance, we might ask test participants to evaluate 20
graphical lineups (400 total plots), even though we expect to reduce the
number to 10 or 15 during the main experiment. We test the application
in individual or focus group sessions, often using graduate students,
colleagues, social media acquaintances, and conscripted family members.
After these participants complete the study, we ask questions about
fatigue to determine what range of trials per participant is reasonable.
At the end of our initial experiment tests, we have enough information
to determine the basic parameters of the experimental design (e.g.~how
many blocks in an incomplete block design can we have with the factors
under investigation). The number of trials a single participant can
complete without excessive fatigue impacts the number of blocks and the
strategy by which we allocate trials to each participant.

In addition, we must consider how long participants take to complete the
required number of trials. Completion time is used to determine
participant compensation (if using a participant recruiting platform).
Ethics boards and some recruitment platforms require that participants
are paid a reasonable wage for their time (currently, around \$15 US per
hour), and platforms may ask for median completion time and
automatically reject submissions from participants who are too far under
or over the specified time limits; they may also require additional
participant payments if the median time estimate is too far below the
actual average completion time during the experiment. Platforms may also
calculate fees based on both the participant payment and number of
participants recruited, with additional fees to recruit
e.g.~demographically representative samples; as a result, it can be
advantageous to balance cognitive load concerns with the fee structure
used by the selected recruitment platform.

The findings from the initial test of the experimental procedure are
then used to revise the data collection procedure in preparation for one
or more pilot studies. It is important to ensure that the software
platform, trial allocation, and other components of the experiment are
functioning as desired before a formal pilot study is conducted. In some
cases, the pilot study is as simple as a ``soft launch'' of the main
experiment, where the total number of trials is pre-specified and only a
few trials are released initially to ensure that data collection works
as expected. In others, the pilot study is conducted first, and results
from that study are used to determine the sample size for the main
experiment. At this point, data collection, analysis, and reporting
proceed much as in any other experiment.

@@

Once the graphical task has been identified, it is necessary to develop
a model which can be used to explore the graphical features of interest
in a precise manner. This is the single longest part of the entire
experimental design and execution process, in part because choosing a
model that replicates important visual features of the data is extremely
complex (Cook et al., 2021; Hullman \& Gelman, 2021; Vanderplas, 2021).

There are two main options when developing a statistical model for
graphical testing: start with a large data set and sample from that data
set (Hofmann et al., 2012), or start from a model and sample data from
that model generating process (Robinson, 2022; Vanderplas \& Hofmann,
2015, 2017). This decision is largely determined by the availability of
a large data set containing the requisite features of interest and the
qualities being manipulated in the experiment. For instance, Hofmann et
al. (2012) used samples of different sizes from a pre-existing data set
to manipulate the amount of signal in each comparison; with a small
sample, there is less signal and the same amount of noise, making the
true plot harder to spot. In many situations, though, a convenient data
set with the right properties is harder to acquire, and it becomes
necessary to develop a sampling model to generate data for user
evaluation.

@@

The choice of the parameter space used in testing is crucial to gain
insight from a study without putting too much burden on participants
with overlong studies. Choosing an appropriate space for testing
parameters is a well-known problem in psychometric testing: the space
considered should cover the area between `only some activation' to
`almost full activation' of an appropriate psychometric function (Schütt
et al., 2016; Valentin et al., 2024). When testing charts, visual
assessment is obviously key, but researchers can make use of statistical
indices related to the testing condition to narrow the parameter space
to a reasonable and efficient subset from which maximal information can
be acquired.\\
These statistical indices may also serve as quantitative proxies for the
difficulty of the visual task. To identify a statistical proxy for
visual difficulty that may help with narrowing the parameter space, it
can be useful to consider numerical measures used to estimate the same
types of visual information that will be assessed in the experiment. For
instance, we have used

@@

Once an appropriate set of parameters are identified using the numerical
screening method, it is important to calibrate these parameter
selections visually. No numerical statistic is a perfect measure of what
we actually see: at best, they are approximations of what we might
potentially see. We have found it to be useful to have one experimenter
calibrate the model parameters at a gross level, and then have another
experimenter narrow in on the parameters which are visually reasonable
within the selected range. Then, both examiners visually inspect a large
number of plots generated using those parameters to get a sense for how
difficult the task at hand is (this strategy is also described by Lu et
al. (2022)). At some point, all experimenters become so visually
saturated with the nuances of the data generating mechanism that it may
become necessary to ``sanity check'' the protocol with family members,
friends, and colleagues. These informal focus groups provide extremely
useful feedback and can help to counteract the visual saturation of
being immersed in the design of a visualization experiment for months at
a time.

@@

We cannot overstate the importance of visual assessment of your model
stimuli, preferably with fresh eyes. We highly recommend performing
several rounds of think-aloud pilot testing (focus) before deploying an
experiment. In support of this assessment, we offer up a cautionary tale
of our own experience: that of Vanderplas \& Hofmann (2017), where we
designed an experiment to test which plot aesthetics promoted discovery
of linear trends and/or clusters.

The experiment was a 2x3x3 factorial exploration of three data
generating parameters, with 3 replicates at each parameter combination
(54 data sets) and 10 aesthetic combinations (for a total of 540
lineups). Each lineup had 20 different sub-panels, so we should have
carefully visually inspected some 10,800 different panels. As is evident
from the fact that we're telling this story as a cautionary tale, we
missed a critical problem with our data-generating mechanism: when
clusters were assigned to randomly generated data after the fact, we
didn't control the cluster size, leading to clusters of one or two
points in relatively few sub-panels. This became particularly noticeable
when bounding ellipses were added to the plot, as the method used to
generate those ellipses required at least 3 points in the cluster. The
missing boundary ellipse in the corresponding sub-panels escaped our
notice during the stimuli proof-reading phase of the experiment, but did
not escape the notice of our participants, who only needed to examine
about 10 lineups each (around 200 panels). An example of one of the
problematic lineups is shown in Figure~\ref{fig-lineup-problems}: many
participants selected panel 16 because of the missing ellipse; not a
wrong choice, but certainly not the effect we intended to test.

@@

\section{Protocol Development}\label{sec-exp-dev}

It would be difficult to develop a full data generating model without
some idea of the experimental protocol: the basic equipment required for
the experiment, some idea of what questions users will answer, where and
how data will be collecteed, and so on. These experimental design
factors are fairly natural for scientists to accumulate over the course
of imagining and planning an experiment. When conducting graphical
tests, however, there are additional considerations beyond those taught
in a standard experimental design course. Experimenters must carefully
consider how much information participants should have about the
experiment, what platform to use to recruit participants, and the
experimental infrastructure underlying data collection. In addition to
our brief overview of considerations in our research, Kochari (2019) has
many helpful suggestions for conducting web-based cognitive and
perceptual studies that also apply to statistical graphics experiments.

Of course, more standard statistical design considerations, like
blocking, randomization, sample size, and analysis methodology are also
important; we will discuss these briefly in
Section~\ref{sec-exp-design}. Here, we focus primarily on the experiment
development process, with critical decisions in a roughly temporal
order.

\subsection{Infrastructure}\label{infrastructure}

We have conducted visualization experiments using a wide variety of
tools: custom web servers running interactive, PHP-based forms, generic
web survey platforms (Qualtrics, Google Forms) for static graphics, and
Shiny applications that control every part of the experiment interface
(instructions, generating completion codes for participant payment,
rendering interactive graphics, and generating fully randomized data for
each participant). In our research, Shiny has provided the right balance
between control over the experimental setting, procedure, etc. and the
intricate details of web server administration and management, however,
this balance is likely different for every lab and potentially for every
experiment.

When using Shiny to collect data, we store participant response data and
experiment parameters using SQLite tables, which are automatically
synchronized with cloud storage and tracked with version control. This
ensures that we have incremental records of the tables during data
collection, and that all data is stored across multiple locations,
guarding against hardware failures. In some studies, data is unique to
each participant; in these cases, we highly recommend saving all
generated data to a database as well, so that it is possible to go back
and examine responses alongside the data used to generate the graphical
stimuli. Hard drive space is extremely cheap relative to almost any
other cost in an experiment; saving all of the data is a sensible
measure.

\subsection{Participant Recruitment}\label{participant-recruitment}

There are several different, commonly used methods for recruiting
participants for visualization experiments and cognitive experiments
more broadly. The selection of participant recruitment method depends on
the infrastructure which will be used in the experiment, but the choice
of participant pool may be more critical to the experiment results than
the infrastructure and modality (Uittenhove et al., 2023).

We have used each of the following strategies in experiments:

\begin{itemize}
\item
  Validated, representative panels of participants offered by
  specialized polling groups. In the US, this includes
  \href{https://amerispeak.norc.org/us/en/amerispeak/about-amerispeak/panel-design.html}{NORC's
  AmeriSpeak} and
  \href{https://www.gallup.com/analytics/318911/us-social-research.aspx}{Gallup
  Panels}
\item
  Recruitment platforms designed for crowdsourcing research tasks,
  including \href{https://www.mturk.com/}{Amazon Mechanical Turk} and
  \href{https://www.prolific.com/}{Prolific}. Researchers may be able to
  obtain representative samples measured against census data along
  variables such as race, age, and sex, but participants are often more
  technologically sophisticated and educated than the general population
  on these platforms.
\item
  In-person studies. In person participants are commonly recruited from
  undergraduate students, but in some experiments it may be preferable
  to recruit participants from the local community. Students are often
  convenient for academic studies, as participation in experiments is
  often a component of introductory course experiential learning
  activities, as in Vanderplas et al. (2024). While the data generated
  from required classroom experiments may have higher variability, and
  it may be hard to generalize findings beyond undergraduate students,
  these studies can be conducted much more cheaply than studies
  conducted through online platforms or panels. Recruiting participants
  from the community for in-person research is also viable, but can be
  much more complicated, however, some topics, including research
  involving persons with specific disabilities, the elderly, or
  subject-matter experts may require recruiting participants outside the
  university.
\item
  Social media and email-based recruitment. Researchers may post
  directly to social media sites such as Twitter/X, Mastodon, and
  BlueSky, forums and discussion sites such as Reddit, or general email
  lists that may be purchased from universities and other marketing
  organizations. If the goal is to recruit a specific population, such
  as meterologists or forensic examiners, forums and email lists may be
  an extremely effective way to recruit participants. We have used the
  \href{https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/}{SampleSize
  subreddit} successfully and obtained participants that, while younger
  and more highly educated than those recruited from a platform, had
  similar results on visual estimation tasks (Vanderplas et al., 2019).
  When using samples obtained from acquaintances, researchers should try
  to obtain more diversity than was present in historical papers, such
  as Cleveland \& McGill (1984), where the authors sampled their
  colleagues (and their wives).
\end{itemize}

Rice et al. (2024) recruited participants using nationally
representative panel samples and found that conclusions from fully
representative samples of the population can be very different from
volunteer samples recruited using other services, as many people are
unmotivated to engage with charts, don't know how to read charts, or
impose pre-formed conclusions onto visual displays. While using
representative panel sampling products offered by organizations such as
NORC or Gallup is more expensive than samples from Amazon MTurk or
Prolific, these results may imply that results from less representative
participant recruitment methods may not generalize to the broader
population. People who participate in studies via online platforms are
more technologically sophisticated and educated than the general
population, and this bias may significantly impact the conclusions.
However, it is also reasonable to argue that higher education might make
someone more likely to use charts and data for decision-making.\\
Researchers designing a study should carefully consider whether the goal
of the study is to generalize results to the adult population or to a
subset of that population who make decisions based on data.

If data will be collected online, then we recommend considering the pros
and cons of panel-based sampling methods and online recruitment
platforms. While Amazon MTurk was once the only large platform for this
type of research (Heer \& Bostock, 2010), many researchers now prefer
Prolific because its participants appear to be more attentive to tasks
(Albert \& Smilek, 2023; Peer et al., 2022). Prolific's focus on
academic research rather than developing training data for machine
learning and AI means that its policies are tailored for these types of
projects and its users may be more interested in science than those on
other platforms. Experimenters should consider available options,
compare pricing structures (as these vary widely), and consider whether
add-on fees for e.g.~demographically representative samples are worth
the additional cost. It is also important to ensure that the platform
supports the type of user engagement required for the study. Online
recruitment platforms are more flexible than many panel survey platforms
and allow experimenters to use a much wider range of stimuli and
experimental designs, but this sacrifices some ability to generalize
results to a wider population.

@@

It can be extremely helpful to include ``practice'' demonstrations of
the task to show the basic process, logic, and reasoning. While it is
tempting to make these tasks fully representative of the type of
judgement which will be required of participants, practice tasks which
are too close to the experimental task may bias participants. We have
found that it works well to have a relatively easy practice task which
utilizes a slightly different type of plot and/or type of data than what
will be tested in the experiment. In cases which require interactivity,
gif animations of the task being carried out are useful, as are
additional visual cues, such as the yellow box used in the `You Draw It'
task\footnote{See a gif of testing with `You Draw It'
  \href{https://i.imgur.com/GM5YSen.gif}{here}} to indicate that there
were points which were not completed. Demonstrations can reduce
cognitive load, but it is often critical not to prime participants to
focus on the specific effect manipulated during the experiment. Finding
the right examples and instructions to use in an experiment is a
delicate process: priming participants can reduce the generalizability
or relevance of the experiment, but if participants are confused about
what question to answer or how to complete the task, response
variability and participant cognitive load may be too high to detect an
effect. Pilot testing with a small group of volunteers who think aloud
while completing the experiment can be extremely helpful in identifying
problems with participant instructions. However, it is important that
some pilot participants have lower levels of mathematical and
statistical training that are similar to the participant population. One
technique that we have used is to wrap pilot testing into a presentation
about the experiment to a general audience, perhaps as part of an
undergraduate course or undergraduate research experience recruitment
activity. The experiential component increases participant understanding
of the research before the presenter explains the scientific goals
behind the project, and it is often easy to get feedback on the
experiment at the same time.

\subsection{Attention Checks and Answer
Validation}\label{attention-checks-and-answer-validation}
